Testing gene methylation enrichment analysis approaches using simulated data

### 2023-10-14 figs/GSE158422_simulate.Rmd
started work on simulation

---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

Here we are splitting the dataset and testing concordance between LA and AL approaches.

@@

@@

>>>>

---
title: "Testing gene methylation enrichment analysis approaches using simulated data"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

In this report we establish a new method for generating simulated data with known ground truth.
This will be used to test different gene methylation enrichment approaches systematically.

The general steps are:

1. Import GSE158422 data corresponding to control (non-tumour tissue).

2. From the 37 samples, create two groups of 18 samples.
One of these will be considered "control" and the other "case".

3. Some gene sets will be selected to be differentially methylated.
A requirement is that there will be no overlap between them.
Half of these will be hypermethylated and the others will be hypomethylated.

4. The changes will be incorporated into the "case" samples.

5. The enrichment analysis will be conducted.

6. The accuracy will be calculated.

@@

# Reactome

@@

Select nonoverlapping gene sets.

```{r,gset_selection}

gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))

gset_selection <- function(genesetdatabase,seed,gene_catalog) {
  set.seed(seed)
  gsets <- gs_symbols[sample(1:length(genesetdatabase))]
  genelist=NULL
  genesets=NULL
  for ( i in 1:length(gsets) ) {
    gs <- gsets[i]
    inx <- length(intersect(unlist(gs),genelist))
    num <- length(intersect(unlist(gs),gene_catalog))
    if ( inx == 0 & num >= 10 ) {
      genesets <- c(genesets,gs)
      genelist <- c(unname(unlist(gs)),genelist)
    }
  }
  return(genesets)
}

gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=100,gene_catalog=gene_catalog)

# Incorporate changes

TODO: to incorporate changes to case samples.

Need to figure out what magnitude to change.
Will refer to the cancer/normal comparison.

Select genes and probes to alter.

```{r,incorp}

seed=100

# select gene sets to alter
gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=seed,gene_catalog=gene_catalog)

# select normal samples only
normal <- mval[,(1:(ncol(mval)/2)*2)]

# divide sets between hyper and hypomethylated
nset <- floor(length(gset_mod)/2)
set.seed(seed) ; gtup <-sample(gset_mod,48)
set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
gup <- unname(unlist(gtup))
gdn <- unname(unlist(gtdn))

# divide samples between ctrl and case
set.seed(seed) ; ctrl <- sample(1:ncol(normal),18)
set.seed(seed) ; case <- sample(setdiff(1:37,ctrl),18)
mval_ctrl <- mval[,ctrl]
mval_case <- mval[,case]

# select probes
